---
title: "Simulation Deep Dive"
author: "Paul Harmon"
date: "10/13/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

source("OutlierCompNew.R")
library(ggplot2)
library(readr)
library(tibble)
library(dplyr)
library(tidyr)
library(janitor)
library(magrittr)
library(knitr)
library(Rtsne)
library(robustHD)
library(GGally)
library(stringr)
library(plotly)
library(R.utils)
library(vegan)
library(mnormt)
library(MASS)



correctInfluence = function(result_df){
  tmp <- tail(result_df,1)
  det_rate <- length(which(tmp$PValues <= 0.05))/nrow(tmp)
  return(det_rate)}

# Correctly Identified Non-Influential
#head(df3list[[1]], 60)

correctNonInfluence = function(result_df){
  tmp <- head(result_df,60)
  det_rate <- length(which(tmp$PValues > 0.05))/nrow(tmp)
  return(det_rate)}


spuriousInfluence = function(result_df){
  tmp <- result_df
  det_rate <- length(which(result_df$PValues <= 0.05))/nrow(tmp)
  return(det_rate)}

#then the non-influential 
meantest = function(df){
  mean(df$PValues)
}

```

# Intro

This document shows some of the results from a handful of different simulations. In general, we consider a situation where the data are arranged in 3 groups of varying width and variability. Each simulation fixes the sample size within groups and all groups are simulated from multivariate normals with 10 features. Additionally, only a SINGLE influential point is considered at present. 

In most scenarios, 61 observations are used - except when no outliers are present. 

Simulation scenarios: 

+ Influential point "bounces around" within range of 3 groups
+ Influential point is "smaller" ~ 3 SD below average
+ Influential point is "larger" ~ 3 SD above average
+ No outliers (only 60 observations)

Pending Scenarios:

+ Wider range of influential points
+ Wider spread of groups
+ More dimensions than 10
+ Influence on only a subset of features


# Multiple Groups - Uncorrelated

## Data


The first section considers an influential point that bounces around within the variation of the three groups.  The means are different across the groups but there is no correlation between the 10 features simulated. 

```{r}
g1 <- rmnorm(n = 10, mean = rep(20, 20), diag(35, 20)) %>%
    t() %>%
    data.frame()
  g2 <- rmnorm(n = 10, mean = rep(40, 20), diag(35, 20)) %>%
    t() %>%
    data.frame()
  g3 <- rmnorm(n = 10, mean = rep(60, 20), diag(35, 20)) %>%
    t() %>%
    data.frame()
  
  
  yes_structure <- rbind(g1, g2, g3) %>%
    data.frame() %>%
    mutate(Group = c(rep("1", 20), rep("2", 20), rep("3", 20)))
  
  yes_structure_scale <- lapply(yes_structure[,1:10], scale) %>% as_tibble()
  
  ### adds outliers - but not as a single group. Here we have an outlier that includes a really big one, a really small observation, and a very abnormal one
  #add_outliers <- rbind(rnorm(10, 3, .5), rnorm(10, -3, .5), rnorm(10, c(rep(-2,3),rep(0,3),rep(2,3),0), .5)) %>% as_tibble()
  
  add_outliers <- rbind(rnorm(10, 0, 1)) %>% as_tibble()
  
  names(add_outliers) <- names(yes_structure_scale)
  
  yes_structure_final <- bind_rows(yes_structure_scale, add_outliers) 
  yes_structure_final$Group <- c(yes_structure$Group, rep("Outlier",1))
  
  
  ##### Visualize the data and store in a list
  datplot <- yes_structure_final %>% mutate(Id = 1:nrow(yes_structure_final), Alpha = ifelse(Group %in% 'Outlier', 1,0.8)) %>%  pivot_longer(1:10) %>% ggplot(aes(name, value, color = Group, group = Id, alpha = Alpha)) + geom_line() + geom_point() + ggtitle("Simulated Data with Outliers Added In") + guides(alpha = FALSE)
datplot
```

## Results

```{r}
mds1 <- readRDS("2021-10-16_simulation1_mds_list.RDS")
tsne1 <- readRDS("2021-10-16_simulation1_tsne_list.RDS")



#### Generate a table of values to assess efficacy of detections -----------------#
df2list <- tsne1
df3list  <- mds1
#note - these will give values per run - so we need to take a look across the simulations

#t-SNE
ci1 <- sapply(df2list, correctInfluence)
cni1 <- sapply(df2list, correctNonInfluence)


#MDS
ci2 <- sapply(df3list, correctInfluence)
cni2 <- sapply(df3list, correctNonInfluence)


#mean detection rates
Category <- c("Correctly Identified Influential", "Non-Influential")
TSNE <- c(mean(ci1), mean(cni1))
MDS <- c(mean(ci2), round(mean(cni2),3))
temp <- cbind(Category,TSNE, MDS)

kable(temp)


```

# Larger - Uncorrelated

In this setting, we examine a case where the influential point is larger than the rest of the data. The data are still simulated in multiple groups. 

## Data 

```{r}
 g1 <- rmnorm(n = 10, mean = rep(20, 20), diag(35, 20)) %>%
    t() %>%
    data.frame()
  g2 <- rmnorm(n = 10, mean = rep(40, 20), diag(35, 20)) %>%
    t() %>%
    data.frame()
  g3 <- rmnorm(n = 10, mean = rep(60, 20), diag(35, 20)) %>%
    t() %>%
    data.frame()
  
  
  yes_structure <- rbind(g1, g2, g3) %>%
    data.frame() %>%
    mutate(Group = c(rep("1", 20), rep("2", 20), rep("3", 20)))
  
  yes_structure_scale <- lapply(yes_structure[,1:10], scale) %>% as_tibble()
  
  ### adds outliers - but not as a single group. Here we have an outlier that includes a really big one, a really small observation, and a very abnormal one
  #add_outliers <- rbind(rnorm(10, 3, .5), rnorm(10, -3, .5), rnorm(10, c(rep(-2,3),rep(0,3),rep(2,3),0), .5)) %>% as_tibble()
  
  add_outliers <- rbind(rnorm(10, 3, .5)) %>% as_tibble()
  
  names(add_outliers) <- names(yes_structure_scale)
  
  yes_structure_final <- bind_rows(yes_structure_scale, add_outliers) 
  yes_structure_final$Group <- c(yes_structure$Group, rep("Outlier",1))
  
  
  ##### Visualize the data and store in a list
  datplot <- yes_structure_final %>% mutate(Id = 1:nrow(yes_structure_final), Alpha = ifelse(Group %in% 'Outlier', 1,0.8)) %>%  pivot_longer(1:10) %>% ggplot(aes(name, value, color = Group, group = Id, alpha = Alpha)) + geom_line() + geom_point() + ggtitle("Simulated Data with Outliers Added In") + guides(alpha = FALSE)
datplot
```


## Results

```{r}
mds1 <- readRDS("2021-10-17_simulationlarger_mds_list.RDS")
tsne1 <- readRDS("2021-10-17_simulationlarger_tsne_list.RDS")


#### Generate a table of values to assess efficacy of detections -----------------#
df2list <- tsne1
df3list  <- mds1
#note - these will give values per run - so we need to take a look across the simulations

#t-SNE
ci1 <- sapply(df2list, correctInfluence)
cni1 <- sapply(df2list, correctNonInfluence)


#MDS
ci2 <- sapply(df3list, correctInfluence)
cni2 <- sapply(df3list, correctNonInfluence)


#mean detection rates
Category <- c("Correctly Identified Influential", "Non-Influential")
TSNE <- c(mean(ci1), round(mean(cni1),3))
MDS <- c(mean(ci2), round(mean(cni2),3))
temp <- cbind(Category,TSNE, MDS)

kable(temp)


```


# Smaller - Uncorrelated

This setting should be quite similar to the results that we see in the larger influential point as the point is simply "flipped" to be several standard deviations smaller across the 10 features in the data. 

## Data 

```{r}
 g1 <- rmnorm(n = 10, mean = rep(20, 20), diag(35, 20)) %>%
    t() %>%
    data.frame()
  g2 <- rmnorm(n = 10, mean = rep(40, 20), diag(35, 20)) %>%
    t() %>%
    data.frame()
  g3 <- rmnorm(n = 10, mean = rep(60, 20), diag(35, 20)) %>%
    t() %>%
    data.frame()
  
  
  yes_structure <- rbind(g1, g2, g3) %>%
    data.frame() %>%
    mutate(Group = c(rep("1", 20), rep("2", 20), rep("3", 20)))
  
  yes_structure_scale <- lapply(yes_structure[,1:10], scale) %>% as_tibble()
  
  ### adds outliers - but not as a single group. Here we have an outlier that includes a really big one, a really small observation, and a very abnormal one
  #add_outliers <- rbind(rnorm(10, 3, .5), rnorm(10, -3, .5), rnorm(10, c(rep(-2,3),rep(0,3),rep(2,3),0), .5)) %>% as_tibble()
  
  add_outliers <- rbind(rnorm(10, -3, .5)) %>% as_tibble()
  
  names(add_outliers) <- names(yes_structure_scale)
  
  yes_structure_final <- bind_rows(yes_structure_scale, add_outliers) 
  yes_structure_final$Group <- c(yes_structure$Group, rep("Outlier",1))
  
  
  ##### Visualize the data and store in a list
  datplot <- yes_structure_final %>% mutate(Id = 1:nrow(yes_structure_final), Alpha = ifelse(Group %in% 'Outlier', 1,0.8)) %>%  pivot_longer(1:10) %>% ggplot(aes(name, value, color = Group, group = Id, alpha = Alpha)) + geom_line() + geom_point() + ggtitle("Simulated Data with Outliers Added In") + guides(alpha = FALSE)
datplot
  
```


## Results
```{r}
mds1 <- readRDS("2021-10-18_simulationsmaller_mds_list.RDS")
tsne1 <- readRDS("2021-10-18_simulationsmaller_tsne_list.RDS")



#### Generate a table of values to assess efficacy of detections -----------------#
df2list <- tsne1
df3list  <- mds1
#note - these will give values per run - so we need to take a look across the simulations

#t-SNE
ci1 <- sapply(df2list, correctInfluence)
cni1 <- sapply(df2list, correctNonInfluence)


#MDS
ci2 <- sapply(df3list, correctInfluence)
cni2 <- sapply(df3list, correctNonInfluence)

#mean detection rates
Category <- c("Correctly Identified Influential", "Non-Influential")
TSNE <- c(mean(ci1), round(mean(cni1),3))
MDS <- c(mean(ci2), round(mean(cni2),3))
temp <- cbind(Category,TSNE, MDS)

kable(temp)



```



# No Outliers - Uncorrelated Data 

## Data 
```{r}
 g1 <- rmnorm(n = 10, mean = rep(20, 20), diag(35, 20)) %>%
    t() %>%
    data.frame()
  g2 <- rmnorm(n = 10, mean = rep(40, 20), diag(35, 20)) %>%
    t() %>%
    data.frame()
  g3 <- rmnorm(n = 10, mean = rep(60, 20), diag(35, 20)) %>%
    t() %>%
    data.frame()
  
  
  yes_structure <- rbind(g1, g2, g3) %>%
    data.frame() %>%
    mutate(Group = c(rep("1", 20), rep("2", 20), rep("3", 20)))
  
  yes_structure_scale <- lapply(yes_structure[,1:10], scale) %>% as_tibble()
  
  ### No outliers - only 60 observations in this dataset 
  
  yes_structure_final <- yes_structure_scale
  yes_structure_final$Group <- c(yes_structure$Group)
  
  
  ##### Visualize the data and store in a list
  datplot <- yes_structure_final %>% mutate(Id = 1:nrow(yes_structure_final), Alpha = ifelse(Group %in% 'Outlier', 1,0.8)) %>%  pivot_longer(1:10) %>% ggplot(aes(name, value, color = Group, group = Id, alpha = Alpha)) + geom_line() + geom_point() + ggtitle("Simulated Data with Outliers Added In") + guides(alpha = FALSE)
datplot
```


## Results
```{r}
#2021-10-18_simulationsmaller_mds_list

mds1 <- readRDS("2021-10-20_simulationNoInfluence_mds_list.RDS")
tsne1 <- readRDS("2021-10-20_simulationNoInfluence_tsne_list.RDS")



#### Generate a table of values to assess efficacy of detections -----------------#
df2list <- tsne1
df3list  <- mds1
#note - these will give values per run - so we need to take a look across the simulations

#t-SNE
ci1 <- sapply(df2list, spuriousInfluence)
cni1 <- 1-ci1


#MDS
ci2 <- sapply(df3list, spuriousInfluence)
cni2 <- 1 - ci2


#mean detection rates
Category <- c("Correctly Identified Influential", "Non-Influential")
TSNE <- c(round(mean(ci1),3), round(mean(cni1),3))
MDS <- c(round(mean(ci2),3), round(mean(cni2),3))
temp <- cbind(Category,TSNE, MDS)

kable(temp)



```




# Correlated Data 


## Data

```{r}
n <- 10
  Sigma <- rWishart(1,n,diag(1,n)) %>% matrix(nrow = n, ncol = 10)
  diag(Sigma) <- 35
  
  
  g1 <- mvrnorm(n = 20, mu = rep(20, 10), Sigma)  %>%
    data.frame()
  g2 <- mvrnorm(n = 20, mu = rep(40, 10), Sigma) %>%
    data.frame()
  g3 <- mvrnorm(n = 20, mu = rep(60, 10), Sigma) %>%
    data.frame()
  
  
  yes_structure <- rbind(g1, g2, g3) %>%
    data.frame() %>%
    mutate(Group = c(rep("1", 20), rep("2", 20), rep("3", 20)))
  
  yes_structure_scale <- lapply(yes_structure[,1:10], scale) %>% as_tibble()
  
  add_outliers <- rbind(rnorm(10, 0, 1)) %>% as_tibble()
  
  names(add_outliers) <- names(yes_structure_scale)
  
  yes_structure_final <- bind_rows(yes_structure_scale, add_outliers) 
  yes_structure_final$Group <- c(yes_structure$Group, rep("Outlier",1))
  
  
  ##### Visualize the data and store in a list
  datplot <- yes_structure_final %>% mutate(Id = 1:nrow(yes_structure_final), Alpha = ifelse(Group %in% 'Outlier', 1,0.8)) %>%  pivot_longer(1:10) %>% ggplot(aes(name, value, color = Group, group = Id, alpha = Alpha)) + geom_line() + geom_point() + ggtitle("Simulated Data with Outliers Added In") + guides(alpha = FALSE)
  datplot
  
```

A heatmap of the variance-covariance matrix used to simulate the data is shown below. The correlations shift a little during the simulations because of the inverse-Wishart distribution used to generate the Sigma matrix at each round, but there are very few extreme correlations present - just some association between the features. 

```{r}
image(Sigma, main = "Variance-Covariance Matrix of Features")

kable(Sigma)
```





## Results 

**Results are pending** - simulation results are based on fewer runs. 

```{r}

tsne1 <- readRDS("2021-10-29_simulationCorrelated_tsne_list2.RDS")
mds1 <- readRDS("2021-10-29_simulationCorrelated_mds_list2.RDS")



#### Generate a table of values to assess efficacy of detections -----------------#
df2list <- tsne1
df3list  <- mds1
#note - these will give values per run - so we need to take a look across the simulations


#t-SNE
ci1 <- sapply(df2list, correctInfluence)
cni1 <- sapply(df2list, correctNonInfluence)


#MDS
ci2 <- sapply(df3list, correctInfluence)
cni2 <- sapply(df3list, correctNonInfluence)


#mean detection rates
Category <- c("Correctly Identified Influential", "Non-Influential")
TSNE <- c(round(mean(ci1),3), round(mean(cni1),3))
MDS <- c(round(mean(ci2),3), round(mean(cni2),3))
temp <- cbind(Category,TSNE, MDS)

kable(temp)

```



# Correlated Single Group

I instead took a different approach to simulating correlated features; although in this case, I did not vary the group means so we end up with a correlated, single group simulation in this setting. 

## Data

The data are shown below: 

```{r}
 ## Simulates from standard normal distribution
  init = matrix(0, nrow = 60, ncol = 10)
  snorms = apply(init, 2, rnorm, n = 60)
  
  varcov = diag(10) #initializes variance covariance
  
  # want to correlate feature 1 with features 2, 3, 4 
  varcov[1,2] <- varcov[2,1] <- 0.9
  varcov[1,3] <- varcov[3,1] <- 0.75
  varcov[1,4] <- varcov[4,1] <- 0.6
  
  yes_structure <- snorms %*% varcov
  
  yes_structure <- yes_structure %>%
    data.frame() %>%
    mutate(Group = c(rep("1", 20), rep("2", 20), rep("3", 20)))
  
  yes_structure_scale <- lapply(yes_structure[,1:10], scale) %>% as_tibble()
  
  add_outliers <- rbind(rnorm(10, 0, 1)) %>% as_tibble()
  
  names(add_outliers) <- names(yes_structure_scale)
  
  yes_structure_final <- bind_rows(yes_structure_scale, add_outliers) 
  yes_structure_final$Group <- c(yes_structure$Group, rep("Outlier",1))
  
  
  ##### Visualize the data and store in a list
  datplot <- yes_structure_final %>% mutate(Id = 1:nrow(yes_structure_final), Alpha = ifelse(Group %in% 'Outlier', 1,0.8)) %>%  pivot_longer(1:10) %>% ggplot(aes(name, value, color = Group, group = Id, alpha = Alpha)) + geom_line() + geom_point() + ggtitle("Simulated Data with Outliers Added In") + guides(alpha = FALSE)
  datplot
```

In this setting, we can see that the correlation structure is present: 

```{r}
library(corrplot)

cor(yes_structure_final[,1:10]) %>% corrplot("number")
```

And the original Sigma matrix is given below: 

```{r}
varcov
image(varcov)
```




## Results 

The results from the simulation are shown below: 

```{r}

tsne1 <- readRDS("2021-10-29_simulationStrongCorrelated_tsne_list2.RDS")
mds1 <- readRDS("2021-10-29_simulationStrongCorrelated_mds_list2.RDS")



#### Generate a table of values to assess efficacy of detections -----------------#
df2list <- tsne1
df3list  <- mds1
#note - these will give values per run - so we need to take a look across the simulations


#t-SNE
ci1 <- sapply(df2list, correctInfluence)
cni1 <- sapply(df2list, correctNonInfluence)


#MDS
ci2 <- sapply(df3list, correctInfluence)
cni2 <- sapply(df3list, correctNonInfluence)


#mean detection rates
Category <- c("Correctly Identified Influential", "Non-Influential")
TSNE <- c(round(mean(ci1),3), round(mean(cni1),3))
MDS <- c(round(mean(ci2),3), round(mean(cni2),3))
temp <- cbind(Category,TSNE, MDS)

kable(temp)




```










# APPENDIX: ONE GROUP, THREE OUTLIERS

## Data

A look at the data, which is simulated from multivariate normal with a single group. 

```{r}
 g1 <- rmnorm(n = 10, mean = rep(40, 60), diag(35, 60)) %>%
    t() %>%
    data.frame()

  yes_structure <- g1 %>%
    data.frame() %>%
    mutate(Group = c(rep("1", 60)))
  
  yes_structure_scale <- lapply(yes_structure[,1:10], scale) %>% as_tibble()
  
  ### adds outliers - but not as a single group. Here we have an outlier that includes a really big one, a really small observation, and a very abnormal one
  add_outliers <- rbind(rnorm(10, 3, .5), rnorm(10, -3, .5), rnorm(10, c(rep(-2,3),rep(0,3),rep(2,3),0), .5)) %>% as_tibble()
  names(add_outliers) <- names(yes_structure_scale)
  
  yes_structure_final <- bind_rows(yes_structure_scale, add_outliers) 
  yes_structure_final$Group <- c(yes_structure$Group, rep("Outlier",3))
  
  
  ##### Visualize the data and store in a list
  datplot <- yes_structure_final %>% mutate(Id = 1:nrow(yes_structure_final), Alpha = ifelse(Group %in% 'Outlier', 1,0.8)) %>%  pivot_longer(1:10) %>% ggplot(aes(name, value, color = Group, group = Id, alpha = Alpha)) + geom_line() + geom_point() + ggtitle("Simulated Data with Outliers Added In") + guides(alpha = FALSE)
datplot
```


## Results 
We see that the proportion of correctly-identified influential points seems to go down as we constrain the shape of the data to a single group. 

```{r}
mds2 <- readRDS("2021-10-08_simulation2_mds_list.RDS")
tsne2 <- readRDS("2021-10-08_simulation2_tsne_list.RDS")

#### Generate a table of values to assess efficacy of detections -----------------#
df2list <- tsne2
df3list  <- mds2
#note - these will give values per run - so we need to take a look across the simulations

#t-SNE
ci1 <- sapply(df2list, correctInfluence)
cni1 <- sapply(df2list, correctNonInfluence)


#MDS
ci2 <- sapply(df3list, correctInfluence)
cni2 <- sapply(df3list, correctNonInfluence)


#mean detection rates
mean(ci1)
mean(cni1)

mean(ci2)
mean(cni2)

```


# Computation Time

```{r}
pt1 <- readRDS('perf_time_1017.RDS')
pt2 <- readRDS('perf_time_cmd_1017.RDS')


unlist(pt1)
#unlist(pt2)


pt1 %>% unlist() %>% as_tibble() %>% ggplot(aes(value)) + geom_histogram(bins = 10) + xlab("Minutes") + ggtitle("Computation Time - TSNE Runs") + theme_bw()


```
















